---
title: "Spatial Data Wrangling in R"
author: "Hannah M Carroll"
output: html_notebook
---

Most of my work involves spatial analysis of some kind and a ton of mapping

- Problem: I hate ArcGIS (a lot)

- Solution: Everything is better in R

---

There are many ways to do these operations. I will go through what I find to be handy.


```{r}

# Set your working directory to the source file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

```


Packages you will need:

```{r}

# List of the packages you need
packages <- c("ggmap", "sp", "rgdal", "rgeos", "maptools", "viridis", "ggplot2", "maps", "raster", "spatialEco", "dplyr") 

# Which ones are new to your machine?
new.packages <- packages[!(packages %in% installed.packages()[,"Package"])]

# If they're new, install them
if(length(new.packages)) install.packages(new.packages)

# Now load all required packages into memory
lapply(packages, library, character.only = TRUE)
```

---

The first step: Get a shapefile

You can download a shapefile directly from a website or ftp

```{r}
if (!file.exists("./ia_eco_l4.shp")){
  download.file(file.path('ftp://newftp.epa.gov/EPADataCommons/ORD/Ecoregions/ia/ia_eco_l4.zip'), 
                f <- tempfile())
  unzip(f, exdir = getwd()) # This unzips the file to your working directory
  rm(f)
}
```
---

Read in with package rgdal
```{r}

# These are the EPA Level 4 ecoregions of Iowa
iowa.regions <- readOGR(dsn=getwd(), layer="ia_eco_l4") 

plot(iowa.regions)

```

---

I'm going to show you a few (of MANY) ways to play with spatial data in R

---

Beginning with ggplot2

```{r}

ggplot() + theme_void() + 
  geom_polygon(data=fortify(iowa.regions), aes(x=long, y=lat, group=group), colour="grey68", fill="grey95")

# Fortify is important. This transforms a spatial data frame into a data frame ggplot2 can handle.

```

---

This map is boring. Flatten the SpatialPolygonsDataFrame to a standard data frame to do more.

```{r}

# Create points for ecoregions
iowa.regions@data$id <- rownames(iowa.regions@data) # Create an id field to join on
iowa.regions.points <- fortify(iowa.regions, region="id") # Extract the points we need from the SPDF

# Join the data with the spatial information by row
iowa.regions.df <- merge.data.frame(iowa.regions.points, iowa.regions@data, by="id")

# Note that fortify is no longer needed because we're working with a data frame
l4map <- ggplot() + theme_void() + 
  geom_polygon(data=iowa.regions.df, aes(x=long, y=lat, group=group, fill=US_L4NAME), color="grey68") +
  scale_fill_viridis_d(name="Level IV Ecoregions", alpha=0.2) + coord_quickmap()

# Adding cord_quickmap gives a basic projection to something plotted from a data frame. It's better (but much slower) to use coord_map()

l4map

```

---

Let's add hillshade from lidar

Use state raster data (the state's link to this is broken as of Wednesday so it's in the GitHub folder)

```{r}
# Read a geotiff with package raster

lidar <- raster("lidar_hs.tif")

plot(lidar) # We'll fix the colors later

```

---

Reprojecting

```{r}

# We're going back to the orginal SPDF for now because we need the projection for the next set of operations.

# Are the projections the same?

identicalCRS(iowa.regions, lidar) # CRS stands for Coordinate Reference System. You can use this check for any two projected spatial objects.

# Nope

```

---

Reprojecting the raster

```{r}

# Reproject the raster to match the state ecoregions

lidar.nad83 <- projectRaster(lidar, crs = crs(iowa.regions))

# Ignore the error. Lidar data will include bad returns that we'll strip out later.

```

---

```{r}

#convert the raster to points
lidar.p <- rasterToPoints(lidar.nad83)

#Make the points a dataframe for ggplot
lidar.df <- data.frame(lidar.p)

#Make appropriate column headings
colnames(lidar.df) <- c("x", "y", "MAP")

# Delete rows with low values for MAP (These are edges and bad returns)
lidar.df <- lidar.df[lidar.df$MAP > 100, ]

```

```{r}

# Make sure we have a good dataset now
summary(lidar.df)

```

---

Make it black and white to lay other colors on top

```{r}

hs.map <- ggplot(lidar.df, aes(x=x, y=y, color=MAP), na.rm=T) + geom_point() + theme_void() +
   scale_color_gradient2(low="black", mid="gray35", high="gray100", midpoint=150, 
                         # Midpoint is going to depend on the landscape. Play with this to get a good contrast.
                       na.value = NA) + guides(color=FALSE)

# Note: it's preferable (faster) to use geom_raster if you don't have two scale fill variables. But we do, so we're using geom_point.

hs.map

```

---

```{r}

# Now lay the ecoregions on top

hs.map + geom_polygon(data=iowa.regions.df, aes(x=long, y=lat, group=group, fill=US_L4NAME), color="grey68", alpha = 0.2) +
  scale_fill_viridis_d(name="Level IV Ecoregions", alpha=0.6) + coord_quickmap()

```



---

Point in poly operations and spatial joins

```{r}
if (!file.exists("./wc2.0_2.5m_prec_05.tif")){
  download.file(file.path('http://biogeo.ucdavis.edu/data/worldclim/v2.0/tif/base/wc2.0_2.5m_prec.zip'), 
                f2 <- tempfile())
  unzip(f2, exdir = getwd()) # This unzips the file to your working directory
  rm(f2)
}

```

```{r}
# Read in May data

# For the full worldclim dataset, use: clim <- getData("worldclim",var="bio",res=2.5)

may.precip <- raster("wc2.0_2.5m_prec_05.tif")

plot(may.precip)

```

Now to clip this to Iowa

```{r}

# Convert a rasterlayer to spdf
may.spdf <- as(may.precip, "SpatialPixelsDataFrame")

```

```{r}

# Transform to match the projection of our regions
may.spdf <- spTransform(may.spdf, CRS(proj4string(iowa.regions)))

```

```{r}

# Clip to Iowa (to make the point in poly operation faster)
ia.precip <- may.spdf[!is.na(over(may.spdf,as(iowa.regions,"SpatialPolygons"))),]

```

```{r}

# Join points (precip) to polygons (regions)
precip.by.region <- point.in.poly(ia.precip, iowa.regions) # from the spatialEco package

```

```{r}

# Flatten to a data frame (or skip and use fortify)
pbyr.df <- as.data.frame(precip.by.region)

# Give it better column names
colnames(pbyr.df)[c(1,22,23)] <- c("mm10", "x", "y")

```

```{r}

# Note that fortify is no longer needed because we're working with a data frame
 ggplot(pbyr.df, aes(x=x, y=y, color=mm10/10)) + theme_void() + geom_point(size=2) +
 geom_polygon(data=iowa.regions.df, aes(x=long, y=lat, group=group), fill=NA, color="grey95") +
  scale_color_viridis(name="Precipitation (mm)", alpha=0.8) + coord_quickmap()

```

```{r}

# See a summary of precip by region

pbyr.df %>%
  group_by(US_L4NAME) %>%
  summarize(mean_may_precip = mean(mm10/10, na.rm = TRUE))

```

---